Universal Characterization of Variability by \\Shannon's $\sum_ip_i\,\textrm{ln}\,p_i$ Formula

### 2019-09-13 Articles/2019-ShannonEntropy/Source/ShannonEntropyIsVariability.tex
Adding introduction and comments (Julian)

\documentclass{article}

%\usepackage{arxiv}

\title{Shannon entropy as the variability of elements within a distribution}

@@

@@

\section{Introduction}

@@

\section{Variability within a distribution}

@@

\section{Units of variability}

@@

\section{Connection to statistical mechanics}

>>>>

\documentclass{article}
\textheight 8.9 in \textwidth 6.5 in \oddsidemargin -10pt \topmargin
-30pt

%\usepackage{arxiv}
%\usepackage[usenames,dvipsnames]{color}

\title{Universal Characterization of Variability by \\Shannon's $\sum_ip_i\,\textrm{ln}\,p_i$ Formula}

@@

\tableofcontents

@@

\section{Introduction\label{int}}

A world without any variety is barely conceivable. It certainly would not admit scientific investigation. One of us recalls but has been unable to relocate a sentence in one of Leibniz's many writings to the effect that reality \emph{is} variety. This is consistent with his view that any contingent thing is defined by a collection of attributes, which necessarily must be distinguishable
at some epistemic level. Paragraph 58 of Leibniz's \emph{Mondadology} asserts that the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''.

Without committing ourselves to Leibniz's overarching principle, our aim in this paper is to draw attention to the specific nature of the variety that can be observed in the universe and the unique suitability of Shannon's entropy (or information) formula $\sum_ip_i\,\textrm{ln}\,p_i$ for the characterization of the variety. Indeed, as we will show in the next section, opportunities for application of Shannon's formula are ubiquitous both on the earth and in the universe at large.

We also wish to comment on the various interpretations of Shannon's formula and attempt to answer this question: what does the formula represent? It is well known that when Shannon recognized the importance of his formula in communication theory he sought von Neumann's advice on the name he should give to the expression he had found. He was told to call it \emph{entropy}, first, because Botlzmann had used an identical formula to define entropy in statistical mechanics and, second, nobody understood entropy and that would provide protection from criticism [exact words to be found]. We believe von Neumann's advice was not totally helpful and has given rise to confusion. Indeed, it is often claimed that Shannon's formula is epistemic in nature and represents ignorance. There certainly are situations in which that is true but equally there are others in which the use of the formula is very appropriate and no ignorance is involved at all. Examples will be given in the next section.

This leads us to propose a different characterization of Shannon's formula that is both more general and more appropriate. Indeed, we will argue that the formula is a measure of the \emph{variability within a distribution of things of a given kind}. Moreover, the formula has a very wide range of applicability because the universe contains many different kinds of things that, through having common attributes, can be recognized as belonging together despite the fact that their attributes are expressed to different degrees or in different ways. To give a first example, the girls in a school can differ in the length and colour of their hair. More examples will follow in the next section.

\section{Comments on Boltzmann's $H$ Theorem}

[This is for you Gabriele]. In Boltzmann's 1872 paper, he introduced the function that later was called the $H$ function, though it seems he meant to call it [upper case] $\eta$ and it was printed as $E$ (which somehow later became $H$). In the paper Boltzmann describes (in $\mu$-space) a gas of monatomic molecules that has already become spatially uniform and also has the same distribution wrt the particles' kinetic energies $x$ at each space point. He therefore considers a spatially constant distribution function $f(x,t)$ and wants to show that whatever its initial value it will tend in accordance with the Boltzmann equation (which he has previously derived and I won't give here) to the Maxwell distribution $f(x,t)=C\sqrt xe^{-hx}$ as $t\rightarrow\infty$. He defines his function $E$ as follows:
\begin{equation}
E=\int_0^\infty f(x,t)\left\{\textrm{log}\left [{f(x,t)\over \sqrt x}\right ]-1\right\}\textrm dx\label{hfun}
\end{equation}
and aims to prove that it cannot increase and attains its least value when $f(x,t)$ becomes the Maxwell distribution. Boltzmann does not give the argument that leads him to the function (\ref{hfun}), but somewhat later in the paper he notes that the $-1$ in the integrand merely reflects the fact that
$$
\int_0^\infty f(x,t)\textrm dx
$$
is the total number of molecules in unit volume, so that 
\begin{equation}
E=\int_0^\infty f(x,t)\textrm{log}\left [{f(x,t)\over \sqrt x}\right ]\textrm dx\label{hfunb}
\end{equation}
also cannot increase.

Boltzmann does not say how he arrived at the function (\ref{hfun}), but its form is obviously dictated by the form of the Maxwell distribution. I'm terrible at equations, but I assume the $\sqrt x$ in the denominator of the integrand in (\ref{hfunb}) is because the usual Shannon expression is `skewed' by the subsidiary Lagrangian multiplier that fixes the total energy to a definite value. Later in the paper, when considering polyatomic molecules, he arrives by means of arguments I have not yet attempted to understand at the familiar $f\, \textrm{log}\,f$ form.

In his very concise 1860 derivation of his energy distribution function, Maxwell commented that

\begin{quote}\small

It appears from this proposition that the velocities are distributed among the particles according  to the same law as the errors are distributed among the observations in the theory of the ``method of least squares''.

\end{quote}\normalsize

In his much more substantial 1867 paper, Maxwell noted that his 1860 derivation could be seen as `precarious' and gave an alternative proof. However, this does not eliminate the mathematical identity of his distribution with the error distribution. Would you agree that this (somewhat indirectly) establishes another example of variablity within a distribution of things [which I think we might use instead of `elements'; `members' (of a set) is another word we might consider]?

\vspace{2in}

TODO: Introduction and Conclusions still need to be done

@@

\section{Variability within a Distribution\label{vwd}}

@@

\section{Units of Variability\label{uv}}

@@

\section{Connection to Statistical Mechanics\label{csm}}
